---
title: "Analyzing and Visualizing Ridership Patterns in Île-de-France"
author: "Group: Melissa MEDJBER, Kamel YENNEK, Elias ACHOUR, Amira HAMDI"
date: "January 2026"
format: html
execute:
  echo: true
  warning: false
  message: false
---


```{r}
#| label: setup
#| include: false
library(tidyverse)
library(lubridate)
library(sf)
```

# 1. Introduction
Public transport planning in Île-de-France relies heavily on detailed knowledge of passenger flows over time and space.
This report analyses smart-card validations on the rail network between 2018 and 2024 to characterise long-term trends, seasonal patterns, and deviations from a “normal” week during holiday periods.

The objectives are threefold:

- To build a reproducible cleaning pipeline that consolidates heterogeneous raw files into a consistent analytical dataset at the stop-area level.

- To perform exploratory data analysis (EDA) on daily, monthly and yearly ridership indicators for the whole network.

- To define and compare reference and holiday weeks, as a basis for a Shiny dashboard that allows interactive exploration of ridership patterns.

# 2. Data and tools
## 2.1 Data sources
The analysis is based on open data provided by Île-de-France Mobilités, including:

- Semester-level rail validation files from 2018 to 2024, with one record per day, stop and ticket category.

- A spatial reference file of “Zones d’arrêt” (stop areas), used to attach geographical coordinates and station names to each stop identifier.

In this report, the study period is restricted to 1 January 2018 – 31 December 2024, and validations are aggregated at the stop-area level defined by the identifier ID_REFA_LDA.

## 2.2 Software and R packages
All computations and figures are produced in R (`version X.Y.Z`) using the following packages:

- ``tidyverse`` for data manipulation (`dplyr, tidyr, readr`) and plotting (`ggplot2`).

- ``lubridate`` for date parsing and handling of temporal variables.

- ``sf`` for reading and transforming the spatial stop-area reference data.

The full analysis is implemented in a reproducible way inside this Quarto document, which can be rendered to PDF or HTML.

# 3. Response to the questions

## 3.1. Data preprocessing and cleaning

### 3.1.1 Reading and concatenating raw yearly files


The raw validation data are provided as semester-based text files for each year between 2018 and 2024.  
For every year, the script imports the two semester files using the function `read.delim()`, but a few files require specific options.  
In particular, the second semester of 2022 uses a semicolon as separator, so we explicitly set `sep = ";"` when reading `validation_2_trim_2022`.  
In 2023, the second semester file is encoded in `UTF-16LE`, which is handled by passing `fileEncoding = "UTF-16LE"` to `read.delim()`.  
Finally, in 2024 the second validation file is named `2024_T3_NB_FER.txt` instead of following the usual `S2` pattern, and is therefore read into the object `validation_2_trim_2024` using its actual file name.

Each semester file is stored in a separate object (e.g. `validation_1_trim_2018`, `validation_2_trim_2018`, …, `validation_2_trim_2024`), and all these objects are then combined into a single list called `liste_dfs`.  
This list structure allows us to apply exactly the same cleaning function to every semester, instead of repeating the same code fourteen times, and ensures that the processing pipeline is compact and reproducible.

```{r}
#| label: import-raw-illustration
#| eval: false
# Example: importing one semester and storing all tables in a list
validation_1_trim_2018 <- read.delim("data/data-rf-2018/2018_S1_NB_FER.txt")
validation_2_trim_2018 <- read.delim("data/data-rf-2018/2018_S2_NB_FER.txt")
liste_dfs <- list(validation_1_trim_2018, validation_2_trim_2018, ...)

```

### 3.1.2 Harmonising column names and types

The historical files do not all share the same schema, so we define a dedicated function ``clean_valid_type()`` that is applied to each element of ``liste_dfs``.
Inside this function, we first standardise the stop-area identifier by renaming any column called ``lda`` or ``ID_ZDC`` into a common field ``ID_REFA_LDA``, which will later be used as the key for spatial joins.

The function then enforces consistent data types across all years: network and stop codes are cast to character (``CODE_STIF_RES, CODE_STIF_ARRET``), the validation count NB_VALD is converted to integer, and the date JOUR is parsed as a Date using flexible formats (``ymd`` or ``dmy``).
Once ``clean_valid_type()`` has been applied to all elements of liste_dfs via purrr::map(), the cleaned tables are concatenated row-wise with bind_rows() to form a single historical dataset ``df`` covering all days, stops and ticket categories between 2018 and 2024.
```{r}
#| label: clean-types-illustration
#| eval: false
# Schematic view of the cleaning function
clean_valid_type <- function(df) {
  df %>%
    rename_with(~ ifelse(. %in% c("lda", "ID_ZDC"), "ID_REFA_LDA", .)) %>%
    mutate(
      CODE_STIF_RES   = as.character(CODE_STIF_RES),
      CODE_STIF_ARRET = as.character(CODE_STIF_ARRET),
      NB_VALD         = as.integer(NB_VALD),
      JOUR            = as.Date(parse_date_time(JOUR, orders = c("ymd", "dmy")))
    )
}
df <- liste_dfs |> purrr::map(clean_valid_type) |> dplyr::bind_rows()

```

### 3.1.3 Handling missing values and inconsistent records

The consolidated dataset df is then examined for data quality issues by counting missing values per column and isolating problematic rows.
We specifically identify records where the stop code ``CODE_STIF_ARRET ``or the network code ``CODE_STIF_RES`` is missing or empty, as well as rows where the stop label ``LIBELLE_ARRET`` is equal to “Inconnu”, which correspond to locations that cannot be reliably mapped to a real station.

To avoid keeping such ambiguous observations, all rows with ``LIBELLE_ARRET == "Inconnu"`` are removed, and the number of deleted lines is reported to quantify the impact of this filter.
We also investigate the distribution of the main ridership variable ``NB_VALD``, verify that there are no negative counts, and drop all rows where ``NB_VALD`` is missing, which correspond to incomplete or non-numeric records in the raw files.

Finally, the time span of the study is enforced by keeping only dates between 1 January 2018 and 31 December 2024 through a simple filter on the JOUR column.
At this stage, ``df`` contains only valid records with non-missing stop and network codes, a non-missing validation count, and dates lying strictly inside the project period.

```{r}
#| label: clean-missing-illustration
#| eval: false
# Schematic example of quality filters
df <- df %>%
  filter(LIBELLE_ARRET != "Inconnu") %>%
  filter(!is.na(NB_VALD)) %>%
  filter(JOUR >= as.Date("2018-01-01"),
         JOUR <= as.Date("2024-12-31"))

```

### 3.1.4 Recoding ticket categories
The original ticket category variable ``CATEGORIE_TITRE`` contains numerous labels with typographical variants and encoding issues (for example “IMAGINE R” vs “Imagine R”, or different spellings for solidarity contracts).
To make the analysis more robust and readable, we recode these raw labels into a limited set of synthetic categories that group similar products together.

In practice, the script uses a ``case_match()`` statement to map several original labels into a single target category.
The main groups are: “Imagine R” (student passes), “Forfait Navigo” (standard season tickets), “Contrat Solidarité Transport” (solidarity fares), “Amethyste”, “Forfaits courts” (short-duration products), “Forfait Gratuité Transport” (FGT), “NON DEFINI”, and a residual group “Autres titres” that absorbs any remaining labels not explicitly listed.

This transformation reduces noise in the typology of tickets and allows comparisons between a small number of meaningful product families instead of many rarely used labels.
```{r}
#| label: recode-tickets-illustration
#| eval: false
# Illustration of the recoding strategy for CATEGORIE_TITRE
df <- df %>%
  mutate(CATEGORIE_TITRE = case_match(
    CATEGORIE_TITRE,
    c("IMAGINE R", "Imagine R") ~ "Imagine R",
    c("NAVIGO", "Forfait Navigo") ~ "Forfait Navigo",
    c("TST", "Contrat Solidarité Transport", "Contrat Solidarit\xe9 Transport") ~
      "Contrat Solidarité Transport",
    c("AMETHYSTE", "Amethyste") ~ "Amethyste",
    c("NAVIGO JOUR", "Forfaits courts") ~ "Forfaits courts",
    "FGT" ~ "Forfait Gratuité Transport",
    c("?", "NON DEFINI") ~ "NON DEFINI",
    .default = "Autres titres"
  ))


```

​

### 3.1.5 Aggregation to stop-area level and spatial join
After cleaning and recoding, the dataset still contains one row per original record, which can correspond to very fine ticket or device details.
For network-level analysis and mapping, we aggregate these records to a more synthetic level: one row per day (``JOUR``), stop-area (``ID_REFA_LDA``) and ticket family (``CATEGORIE_TITRE``), with the total number of validations NB_VALD summed within each group.

This aggregation produces a compact table ``df_agg`` where each row corresponds to the daily ridership of a given stop-area and ticket category.
To enable geographical visualisations, we then join ``df_agg`` with the spatial reference of stop areas, which is stored as a shapefile containing the polygons of “Zones d’arrêt”.

The spatial file is read with the ``sf`` package and transformed to the standard WGS84 coordinate system.

For each zone, we compute its centroid and extract its longitude and latitude, keeping only a few attributes: the stop-area identifier (``ID_REFA_LDA``), the station name (``NOM_GARE``), the municipality (``VILLE``) and the coordinates (``LON, LAT``).
A left join between ``df_agg`` and this cleaned spatial table (``geo_clean``) yields the final analytical dataset ``df_final``, which combines daily ridership with station names and geographic coordinates and is saved as ``data_clean.rds`` for later use in the EDA and the Shiny dashboard.

```{r}
#| label: agg-geo-illustration
#| eval: false
# Aggregation by day, stop-area and ticket family
df_agg <- df %>%
  group_by(JOUR, ID_REFA_LDA, CATEGORIE_TITRE) %>%
  summarise(NB_VALD = sum(NB_VALD, na.rm = TRUE), .groups = "drop")

# Spatial reference of stop areas (simplified workflow)
geo_shp <- st_read("data/geo/PL_ZDL_R_07-01-2026.shp")
geo_clean <- geo_shp %>%
  st_transform(4326) %>%
  st_centroid() %>%
  mutate(
    LON         = st_coordinates(.)[, 1],
    LAT         = st_coordinates(.)[, 2],
    ID_REFA_LDA = as.numeric(idrefa_lda)
  ) %>%
  st_drop_geometry() %>%
  select(ID_REFA_LDA, NOM_GARE = nom_lda, VILLE = commune, LON, LAT)

df_final <- df_agg %>%
  left_join(geo_clean, by = "ID_REFA_LDA")


```

This final table df_final is the main input for all subsequent exploratory analyses and for the Shiny dashboard, since it combines daily ridership by stop-area with station names and geographic coordinates.
